The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-07-16 VegOzone.tex
Spell checked. Completed Introduction draft. Update abstract.

\begin{abstract}
  As a result of successful air quality regulations, the past two decades saw a steady reduction in episodic high summer peak concentrations of ozone over Europe and North America. At the same time, tropospheric background ozone levels kept increasing in the northern hemisphere until recently. An increased occurrence of persistent heatwaves, as one possible future scenario, generates favorable conditions for the formation of ambient air ozone. Vegetation, highly specialized to sub-arctic climate, is vulnerable to rapid environmental changes inflicted by global warming and might become more susceptible to ozone in the future. Over large parts of Europe the summer 2018 had been extraordinarily hot and dry and caused large wildfires in northern Sweden in particular. It can be regarded as a test case for future scenarios. In both 2018 and 2019, we have monitored ambient air ozone concentrations at NIBIO Environment Centre Svanhovd in Northern Norway. Ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance differed significantly, precipitation varied only to some extend. Coincidentally, we have observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019.

  \emph{This may be due to differences in uptake of ozone ($\mathrm{POD_y}$) between the two years. Since DO3SE can be run on birch/beech and conifers, we would like to investigate whether the critical levels for these tree species were crossed during the two summer seasons. Ozone concentrations for the two missing weeks in July 2018 have been estimated based on the measurements from the Swedish and Finnish ozone measurement stations in the region.}  
\end{abstract}

@@

Tropospheric and in particular ambient air ozone is a highly toxic pollutant to all ecosystems. In continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is cause to early senescence as well as programmed cell death in plants {\bf TODO citations}. On average, a global loss of yield in four major crops (wheat, rice, maize, and soybean) of $\approx 3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Former can proof a vital threat to food security in fast developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where air pollution related to ozone has increased in the past years. At the same time, a successful implementation of air quality regulations reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone itself is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}, but in combination with its damaging effects on global vegetation it might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

Tropospheric ozone is formed in complex photochemical cycles in the presents of nitrogen oxides (\chem{NO_x}) involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as (biogenic) volatile organic substances (VOCs and BVOCs). In the absence of local pollution sources, episodes of high ozone concentrations can be attributed to long-range transport of precursors. Despite a relatively short tropospheric life-time of roughly $22\,\unit{days}$ (a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere) \citep{JGR:Stevenson2005,ACP:Young2013}, ozone itself is subject to advection over long distances. In northern Fennoscandia episodes of especially high concentrations of ambient air ozone in 2003 and 2006 have been traced back to emissions related to forest fires in southern Europe and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Boreal wildfires emit ozone precursors such as \chem{CO} ($6-13\,\unit{\%}$ relative to \chem{CO_2^{tot}}) and \chem{VOCs} ($0.5-1.5\,\unit{\%}$ relative to \chem{CO_2^{tot}}) \citep{AE:Cofer1990}.

Ozone displays an annual cycle with a distinct maximum typically in spring \citep{AE:Monks2000}. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. The cause for this might be a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}.

\emph{Contributor: Ane}\\
As mentioned above, ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.

{\bf TODO:
\begin{itemize}
\item general about ozone $\rightarrow$ focus on surface ozone and in particular precursors with respect to the forest fires (\chem{CO}) and heat stress (\chem{VOCs}).
\item ozone induced damage in vegetation 
\item[$\rightarrow$] Some info on stomatal conductance and its empirical calculation:
  \begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within the range $0-1$. They account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type $k$.
\item metrics of ozone
  \begin{itemize}
  \item[$\rightarrow$] Concentration based
    \begin{equation}
      \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. A typical value for $\mathrm{x}=40\,\unit{ppb}$.
  \item[$\rightarrow$] Flux based
    \begin{equation}
      \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$.
  \end{itemize}
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak \citep{AE:Monks2000, AB:Klingberg2009, AE:Rummukainen1996} {\bf TODO: Find newer articles.}
  \end{itemize}
\end{itemize}
}

@@

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}.

@@

{\bf TODO: Future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak) \citep{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007}
  \item[$\rightarrow$] increase in tropospheric background ozone cite WMO?
  \end{itemize}
}

In terms of AOT40, the threshold for protection of vegetation of $3000\,\unit{ppb}$ given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019}. {\bf TODO: come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia. Later supplement figure \ref{fig:fennoscandic_aot40}}

@@

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Swedish (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest \citep{SOU2019}. {\bf TODO: Do we need a corresponding \chem{CO_2} emission here?} However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Appendix Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm}. {\bf TODO: Rewrite how growing season has been calculated.}

@@

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The effective changes from the MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, an assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme have lead to the enhanced performance of latter. The higher spatial resolution of the regional air quality models used in the CAMS Regional Ensemble compared to the global reanalyses does capture more localized depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models with respect to daily and seasonal cycles in ozone {\bf TODO: Any citable source for this assumption?}. Latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results are the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough {\bf TODO: Any citable source for this?}.

>>>>

\begin{abstract}
  As a result of successful air quality regulations, the past two decades saw a steady reduction in episodic high summer peak concentrations of ozone over Europe and North America. At the same time, tropospheric background ozone levels kept increasing in the northern hemisphere until recently. An increased occurrence of persistent heatwaves, as one possible future scenario, generates favorable conditions for the formation of ambient air ozone. Vegetation, highly specialized to sub-arctic climate, is vulnerable to rapid environmental changes inflicted by global warming and might become more susceptible to ozone in the future. Over large parts of Europe the summer 2018 had been extraordinarily hot and dry and caused large wildfires in northern Sweden in particular. This can be regarded as a test case for a future scenario. In both 2018 and 2019, we have monitored ambient air ozone concentrations at NIBIO Environment Centre Svanhovd in Northern Norway. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged significantly from multi annual mean, precipitation varied only to some extend. Coincidentally, we have observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019. We investigate the difference in uptake of ozone using the DO3SE model, with respect to the typical vegetation (beech/birch and conifers) at the location. We assess whether critical levels for these species were breached during the summer seasons. Due to data acquisition problem, ozone concentrations for two weeks in July 2018 were missing from our record and have been reconstructed based on the measurements from the Swedish and Finnish ozone measurement stations in the region.
\end{abstract}

@@

Tropospheric and in particular ambient air ozone is a highly toxic pollutant to all ecosystems. In continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is cause to early senescence as well as programmed cell death in plants {\bf TODO citations}. On average, a global loss of yield in four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Former can proof a vital threat to food security in fast developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone related to air pollution has increased in the past years. At the same time, a successful implementation of air quality regulations reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone in itself is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}. Its damaging effects on global vegetation might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

In the presents of nitrogen oxides (\chem{NO_x}), tropospheric ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs). In the absence of local pollution sources, episodes of high ozone concentrations are often attributed to long-range transport of precursors. In northern Fennoscandia episodes in 2003 and 2006 have been traced back to emissions related to forest fires in southern Europe and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. But also living vegetation emits hydrocarbons in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). Those emissions are associated with, e.g., thermal stress, defense against herbivores, or attraction of pollinators \citep{TPS:Penuelas2003}.

Despite a relatively short tropospheric life-time of roughly $22\,\unit{days}$ ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone itself is subject to advection over long distances.
In the annual cycle a distinct maximum typically occurs in spring followed by lower concentrations throughout the summer probably due to a higher uptake on land vegetation than on bare ground or snow and ice \citep{AE:Monks2000}. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

At the same time, the start of growing season in northern Fennoscandia is shifting to earlier days of the year in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby into close vicinity of the ozone spring peak. This may promote a higher susceptibility of deciduous species to ozone.

Since the industrial revolution, tropospheric background concentrations of ozone in the northern hemisphere have doubled and kept increasing until recently. But long-term observations started before 1950 are scarce \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ACP:Wespes2018,ESA:Gaudel2018}.\\

\emph{Contributor: Ane} {\bf TODO: More detailed ozone damage description?}\\
Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.

The opening of stomata, described through their conductance to gases $g_\mathrm{sto}$, is governed by different environmental factors and is described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$. Stomata are typically considered closed below a light intensity of $50\,\unit{W\,m^{-2}}$.

To measure the risk of damage inflicted to vegetation through ozone, many different metrics have been defined. Two main types can be differentiated, concentration-based (e.g., $\mathrm{AOT_{40}}$) and flux-based (e.g.,$\mathrm{POD_{0.1}}$) metrics. Due to their simplicity, $\mathrm{AOT_{x}}$ are still widely used, despite the actual integrated flux of ozone entering through the stomata is more relevant for an estimate of the actual toxic dose. 
The Accumulated Ozone Exposure over a threshold concentration of $x\,\unit{ppb}$ is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. The integration time varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.
The Phototoxic Ozone Dose (POD) over a threshold y is defined as:
\begin{equation}
  \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$. The stomatal conductance is either measured directly or computed in accordance to Eq.~(\ref{eq:stomatal}).\\

@@

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}.
Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.

@@

In terms of AOT40, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019}. {\bf TODO: come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia. Later supplement figure \ref{fig:fennoscandic_aot40}}

@@

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Swedish (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest \citep{SOU2019}. However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Appendix Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm}.

@@

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The effective changes from the MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, an assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme have lead to the enhanced performance of latter. The higher spatial resolution of the regional air quality models used in the CAMS Regional Ensemble compared to the global reanalyses does capture more localized depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models with respect to daily and seasonal cycles in ozone. Latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results are the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough \citet{ACP:Andersson2017}.
